A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-13 paper/paper.tex
Add text about automation

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{ottalt}
\usepackage{xspace}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

We use the Autosubst 2 framework to specify our syntax in HOAS and
produce Coq syntax files in de Bruijn representation. Additionallly,
Autosubst 2 provides a powerful tactic \texttt{asimpl} which
can be used to prove the equivalence of two terms constructed using
the primitive operators provided by the framework. Instead of
examining carefully the freeness conditions of variables to decide
which rewriting lemma to invoke, reasoning about substitution with
Autosubst 2 boils down to reducing the problem to a goal of the

\section{Discussion}
\label{sec:discuss}

@@

\subsection{Toward Function $\eta$-laws and Decidability of Type Conversion}

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{ottalt}
\usepackage{minted}
\usepackage{xspace}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

We use the Autosubst 2 framework to specify our syntax in HOAS and
produce Coq syntax files in de Bruijn representation. Additionallly,
Autosubst 2 provides a powerful tactic \texttt{asimpl} which
can be used to prove the equivalence of two terms constructed using
the primitive operators provided by the framework. This greatly
simplifies the reasoning about substitution in our development as
almost all substitution related properties are immediately discharged
by \texttt{asimpl} without having to manually prove or even directly invoke any
substitution related lemmas.

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
\begin{minted}{coq}
Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the
graph of the \texttt{tstar} function, which stands for the Takahashi
translation~\citep{takahashi-parallel-reduction}. \texttt{inv:Par} and
\texttt{ctrs:Par} says that the proof involves inverting and
constructing of the derivations of parallel
reduction. The option \texttt{use:Par\_refl,par\_cong} says that the proof would
require the reflexivity and congruence properties of parallel
reduction that we have shown earlier.

The automation here not only gives us a proof that is shorter and more
resiliant to changes in definition, but also gives useful
documentation for readers who wish to understand how the underlying
proof works since the automation tactic is guaranteed to not use
lemmas that are not specified in the \texttt{use} flag.
\section{Discussion}
\label{sec:discuss}

@@

\subsection{Toward Decidability of Type Conversion and $\eta$ laws}
In this section, we sketch out how the logical relation can be extended
show the existence of $\beta\eta$ normal forms for well-typed terms.
In other words, our goal is to show that the following property holds
in our system.
